---
title: "Calculate gene-specific transcription elongation rates using DRB/TTchem-seq2 data"
author: Haiyue Liu
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
    theme: united
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, error = TRUE, cache = TRUE)
knitr::opts_knit$set(root.dir = "/path/to/workdir")
```

```{r libraries}
### Load libraries
suppressPackageStartupMessages({
  library(rtracklayer)
  library(GenomicFeatures)
  library(tibble)
  library(dplyr)
  library(tidyr)
  library(magrittr)
  library(ggplot2)
})
### Define %notin% operator
`%notin%` <- Negate(`%in%`)
```

```{r directories}
bam.dir         <- "/path/to/bam_files/"
output.dir      <- "/path/to/output_files/"
gtf.file        <- "/path/to/annotation.gtf"
sample.info.file <- "/path/to/sample.info.file"   ### tab-delimited file with columns: sample_name, size_factor, sample, rep
count.file      <- "/path/to/count.table.txt"
```

```{r helper_functions}
source("/path/to/help_functions.R")
```

```{r files}
sample.info          <- read.table(sample.info.file, header = TRUE, stringsAsFactors = FALSE)
count.data           <- read.table(count.file, header = TRUE, stringsAsFactors = FALSE)
control.sample.names <- c("DRB_noR_rep1", "DRB_noR_rep2")   ### set to NULL if no control samples
control.sample       <- "DRB_noR"                            ### set to NULL if no control samples
```

## 1. Filter gene annotation

Exclude genes that overlap with other genes on the same strand and genes shorter than 30 kb.

```{r filter_gene_anno}
min.genewidth <- 30000
max.genewidth <- Inf

gene.gr <- filter_overlapping_genes(
  gtf          = gtf.file,
  lower.length = min.genewidth,
  upper.length = max.genewidth,
  count.data   = count.data          ### restricts to expressed genes when provided
)
gene.gr$strand <- strand(gene.gr)
cat("Genes remaining after filtering overlapping genes and applying length limits:", length(gene.gr), "\n")

### Split genes by strand
gene.gr.fwd <- gene.gr[strand(gene.gr) == "+"]
gene.gr.rev <- gene.gr[strand(gene.gr) == "-"]
```

## 2. Count reads at single-base resolution

Using `bamsignals::bamCoverage`, we count reads for each gene from TSS to TES at single-base resolution, then normalize by the spike-in size factor pre-calculated with DESeq2.

Inputs: (1) bam files, (2) sample.info table, (3) gene range object  
Output: spike-in normalized per-base coverage list

```{r count_reads_at_single_bp}
cov.list <- list()

for (sample in sample.info$sample_name) {
  print(sample)

  size.factor  <- sample.info$size_factor[sample.info$sample_name == sample]
  bam.fwd.file <- file.path(bam.dir, paste0(sample, ".fwd.bam"))
  bam.rev.file <- file.path(bam.dir, paste0(sample, ".rev.bam"))

  ### Count read depth on forward and reverse strand genes
  cov.fwd <- bamsignals::bamCoverage(
    bampath     = bam.fwd.file,
    gr          = gene.gr.fwd,
    filteredFlag = 1024,
    paired.end  = "ignore",
    mapqual     = 20,
    verbose     = FALSE
  )
  cov.rev <- bamsignals::bamCoverage(
    bampath     = bam.rev.file,
    gr          = gene.gr.rev,
    filteredFlag = 1024,
    paired.end  = "ignore",
    mapqual     = 20,
    verbose     = FALSE
  )

  ### Extract signal vectors and assign gene IDs
  cov.fwd <- cov.fwd@signals
  cov.rev <- cov.rev@signals
  names(cov.fwd) <- gene.gr.fwd$gene_id
  names(cov.rev) <- gene.gr.rev$gene_id

  ### Combine forward and reverse strand genes
  cov.list[[sample]] <- c(cov.fwd, cov.rev)

  ### Remove genes with zero counts (skip for control samples)
  if (!(sample %in% control.sample.names)) {
    # FIX: lapply returns a list — list > 0 does not give a logical vector for subsetting.
    #      Use sapply to get a named logical vector.
    cov.list[[sample]] <- cov.list[[sample]][sapply(cov.list[[sample]], sum) > 0]
  }

  ### Normalize by spike-in size factor
  cov.list[[sample]] <- lapply(cov.list[[sample]], function(x) { round(x / size.factor, 3) })
}
rm(cov.fwd, cov.rev)
gc()

### Restrict to genes present in all samples
common.genes <- Reduce(intersect, lapply(cov.list, names))
for (sample in names(cov.list)) {
  cov.list[[sample]] <- cov.list[[sample]][common.genes]
}
cat("Counting bp-resolution read depths for", length(cov.list[[1]]), "genes.\n")
```

## 3. Remove lowly expressed genes

Filter out genes with mean coverage < 0.01 in nascent samples. Adjust the threshold based on your data distribution.

```{r remove_lowly_exp_genes}
mean.cov.cutoff <- 0.01

# FIX: pass sample.info explicitly (was previously read from global scope inside the function)
gene.anno <- filter_low_expr_genes(
  cov.list        = cov.list,
  mean.cov.cutoff = mean.cov.cutoff,
  control.samples = control.sample.names,
  sample.info     = sample.info
)
genes.keep <- gene.anno$genes.keep
cat("Genes retained after expression filtering:", length(genes.keep), "\n")

for (sample in names(cov.list)) {
  cov.list[[sample]] <- cov.list[[sample]][genes.keep]
}
```

## 4. Normalize coverage to control samples (optional)

Control samples (DRB not washed out, 4sU incorporated) estimate background signal. DRB release samples are normalized by subtracting the control mean. Skip if no control samples are available.

```{r normalize_to_control}
### Average coverage across replicates
mean.cov.list <- list()
for (sample in unique(sample.info$sample)) {
  print(sample)
  list1 <- cov.list[[ sample.info$sample_name[sample.info$sample == sample & sample.info$rep == "rep1"] ]]
  list2 <- cov.list[[ sample.info$sample_name[sample.info$sample == sample & sample.info$rep == "rep2"] ]]
  for (m in names(list1)) {
    mean.cov.list[[sample]][[m]] <- rowMeans(cbind(list1[[m]], list2[[m]]))
  }
}

if (!is.null(control.sample.names)) {
  ### Subtract control mean coverage from each DRB release sample
  norm.cov.list <- list()
  for (sample in names(mean.cov.list)) {
    if (sample %notin% control.sample) {
      print(sample)
      list.drb  <- mean.cov.list[[sample]]
      list.ctrl <- mean.cov.list[[control.sample]]
      for (m in names(list.drb)) {
        norm.cov.list[[sample]][[m]] <- pmax(0, list.drb[[m]] - list.ctrl[[m]])
      }
    }
  }
} else {
  norm.cov.list <- mean.cov.list
}

rm(cov.list, mean.cov.list)
gc()
```

## 5. Smooth read coverage profiles

Smooth per-gene coverage using `smooth.spline` with `spar = 0.9`, following Gregersen et al. (2020).

```{r smooth_profile}
gene.spar <- 0.9
norm.spline.list <- list()

for (sample in names(norm.cov.list)) {
  print(sample)
  norm.spline.list[[sample]] <- sapply(
    norm.cov.list[[sample]],
    function(x) { pmax(0, round(smooth.spline(seq_along(x), x, spar = gene.spar)$y, 3)) }
  )
}

saveRDS(norm.cov.list,    file.path(output.dir, "norm.cov.list.Rds"))
saveRDS(norm.spline.list, file.path(output.dir, "norm.spline.list.Rds"))
```

## 6. Wave front identification

Identify the leading edge of the elongation wave (furthest distance RNAPII has traveled) for each gene and time point.

```{r find_wave_front}
wf <- sapply(norm.spline.list, function(x) {
  sapply(x, function(i) { wave_front_calling(i) })
}) %>% as.data.frame()

wf[is.na(wf)] <- 0
```

## 7. Elongation rate calculation

Fit a linear model of wave-front distance (kb) ~ DRB release time (min) per gene. Genes with monotonically ordered wave fronts across all four time points use all four points; genes where the 20 min front is at gene end use the first three.

```{r linear_regression}
### Determine wave front ordering per gene
wave.orders <- apply(wf, 1, function(x) { paste0(order(x), collapse = "") })

rate_df <-
  wf %>%
  rownames_to_column("gene_id") %>%
  left_join(
    gene.gr %>%
      as.data.frame() %>%
      mutate(gene_length = width) %>%
      dplyr::select(gene_id, gene_name, gene_type, gene_length),
    by = "gene_id"   # FIX: explicit join key (best practice; avoids ambiguity if column names overlap)
  ) %>%
  pivot_longer(
    cols     = -c(gene_id, gene_type, gene_name, gene_length),
    names_to  = "DRB",
    values_to = "wf"
  ) %>%
  mutate(
    wf   = as.numeric(wf / 1000),
    time = as.numeric(sub("minR", "", sub("DRB_", "", DRB)))
  ) %>%
  left_join(
    as.data.frame(wave.orders) %>% rownames_to_column("gene_id"),
    by = "gene_id"
  )

### Recheck wave front order and assign order category
rate_df %<>%
  group_by(gene_id) %>%
  mutate(order = ifelse(
    (wf[DRB == "DRB_5minR"]  < wf[DRB == "DRB_10minR"]) &
    (wf[DRB == "DRB_10minR"] < wf[DRB == "DRB_15minR"]) &
    (wf[DRB == "DRB_15minR"] < wf[DRB == "DRB_20minR"]) &
    (wf[DRB == "DRB_20minR"] != round(first(gene_length) / 1000, 1)),
    "1234",
    ifelse(
      (wf[DRB == "DRB_5minR"]  < wf[DRB == "DRB_10minR"]) &
      (wf[DRB == "DRB_10minR"] < wf[DRB == "DRB_15minR"]),
      "123", "other"
    )
  )) %>%
  ungroup()

### Fit linear model once per gene and extract slope + R²
# FIX: original code called lm() twice per gene (once for slope, once for R²).
#      Fit once, extract both metrics from the single model object.
rate_df %<>%
  group_by(gene_id) %>%
  mutate(
    fit = list(
      if (first(order) == "1234") {
        lm(wf ~ time)
      } else if (first(order) == "123") {
        lm(wf[DRB != "DRB_20minR"] ~ time[DRB != "DRB_20minR"])
      } else {
        lm(wf ~ time)
      }
    ),
    elongation.rate = fit[[1]]$coef[2],
    adj.r.squared   = summary(fit[[1]])$adj.r.squared
  ) %>%
  dplyr::select(-fit) %>%
  ungroup()
```

```{r save_data}
write.table(
  rate_df,
  file.path(output.dir, "elongation_rate_tb.txt"),
  quote     = FALSE,
  sep       = "\t",
  row.names = FALSE
)
```

```{r session_info}
sessionInfo()
```
