---
title: "Calculate gene-specific transcription elongation rates using DRB/TTchem-seq2 data"
author: Haiyue Liu
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
    theme: united
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, error = TRUE, cache = TRUE)
knitr::opts_knit$set(root.dir = "/path/to/workdir")
```

```{r libraries}
### load libraries
suppressPackageStartupMessages({
  library(rtracklayer)
  library(GenomicFeatures)
  library(tibble)
  library(dplyr)
  library(tidyr)
  library(magrittr)
  library(ggplot2)
})
### define notin  function 
`%notin%` <- Negate(`%in%`)
```


```{r directories}
bam.dir <- "/path/to/bam_files/"
output.dir <- "/path/to/output_files/"
gtf.file <- "/path/to/annotation.gtf"   
sample.info.file <- "/path/to/sample.info.file"        ### table deleminated file containing "sample", "fact_size" columns
count.file <- "/path/to/count.table.txt"
```

```{r helper_functions}
source("/path/to/helper_functions.R")
```

```{r files}
sample.info <- read.table(sample.info.file, header = TRUE)
count.data <- read.table(count.file, header = TRUE)
control.sample.names <- c("DRB_noR_rep1", "DRB_noR_rep2")       ### if no control samples, this can be set as NULL
control.sample <- "DRB_noR"                                     ### if no control samples, this can be set as NULL
```

## 1. Filter gene annotation 

exclude genes that overlap with other genes on the same strand and shorted than 30 kb. 

```{r filter_gene_anno}
min.genewidth <- 30000
max.genewidth <- NULL
gene.gr <- filter_overlapping_genes(gtf = gtf.file,                     
                                    lower.length = min.genewidth,
                                    upper.length = max.genewidth,
                                    count.data = count.data)                  ### if count.data is provided, we only look at genes that are expressed in the dataset
gene.gr$strand <- strand(gene.gr)
cat("The number of genes left after filtering overlapping genes and limiting gene lengths:", length(gene.gr))

### split genes by strand
gene.gr.fwd <- gene.gr[strand(gene.gr) == "+"]
gene.gr.rev <- gene.gr[strand(gene.gr) == "-"]

```

## 2. Count reads at single-base 

Utilizing R bamCoverage function from the bamsignals package, we count the reads for each individual gene from its transcription start site [TSS] to its transcription end site [TES] at single-base resolution. An example of sample.info table can be found in the script folder. Once read coverage is counted, it is normalized by the spike-in size factor that have been pre-calculated using DESeq2. 

Inputs:

(1) read alignments (bam) 
(2) sample.info (text table)
(3) gene range file (gene.gr)

Outputs:

(1) normalized read coverage at single-bp positions for individual genes (list)

```{r count_reads_at_single_bp}
cov.list <- list()
for (sample in sample.info$sample_name){
  print(sample)
  ### Retrieve size factor
  size.factor = sample.info$size_factor[sample.info$sample_name==sample]
  
  bam.fwd.file = file.path(bam.dir, paste0(sample, ".fwd.bam"))
  bam.rev.file = file.path(bam.dir, paste0(sample, ".rev.bam"))
  
  ### Count read depth forward and reverse genes
  cov.fwd <- bamsignals::bamCoverage(bampath = bam.fwd.file, 
                                     gr = gene.gr.fwd,
                                     filteredFlag = 1024,
                                     paired.end = "ignore",
                                     mapqual = 20,
                                     verbose = FALSE)
  
  cov.rev <- bamsignals::bamCoverage(bampath = bam.rev.file,
                                     gr = gene.gr.rev,
                                     filteredFlag = 1024,
                                     paired.end = "ignore",
                                     mapqual = 20,
                                     verbose = FALSE)
  ### Extract values
  cov.fwd <- cov.fwd@signals
  cov.rev <- cov.rev@signals
  
  ### Add gene ids 
  names(cov.fwd) <- gene.gr.fwd$gene_id
  names(cov.rev) <- gene.gr.rev$gene_id
  
  ### Combine genes on the forward and reverse strands
  cov.list[[sample]] <- c(cov.fwd, cov.rev)
  
  ### Remove genes with zero counts
  if(!(sample %in% control.sample.names)){
    cov.list[[sample]] <- cov.list[[sample]][lapply(cov.list[[sample]], sum) > 0]
  }
  ### Normalize by size factors 
  cov.list[[sample]] <- lapply(cov.list[[sample]], function(x){round(x/size.factor, 3)})
}
rm(cov.fwd)
rm(cov.rev)
gc()

### Keep only genes expressed in all samples 
common.genes <- Reduce(intersect, lapply(cov.list, names))
for(sample in names(cov.list)){
  cov.list[[sample]] <- cov.list[[sample]][common.genes]
}
cat("Count bp-resolution read depths for", length(cov.list[[1]]), "genes. \n")
```

## 3. Remove lowly expressed genes

To save time and space for downstream analysis, we only focus on genes with high coverage. In our case, we filter out genes whose mean coverage values are less than 0.01 in the nascent samples. Ideally, you explore the distribution of the read coverage files and define the threshold accordingly. 

```{r remove_lowly_exp_genes}
mean.cov.cutoff <- 0.01

gene.anno <- filter_low_expr_genes(cov.list = cov.list, 
                                    mean.cov.cutoff = mean.cov.cutoff, 
                                    control.samples = control.sample.names)
genes.keep <- gene.anno$genes.keep
cat("Keep ", length(genes.keep), " genes after filtering lowly expressed genes. \n")

for (sample in names(cov.list)){
  cov.list[[sample]] <- cov.list[[sample]][genes.keep]
}
```

## 4. Normalize coverage to control samples (Optional)

We suggest to perform DRB/TTchem-seq2 with control samples in which DRB is not washed out but 4sU is incorporated for the same time as the other samples to estimate the background signals. 

If you do have control samples, the DRB release samples are normalized to controls samples. If not, this step can be ignored. 

```{r normalize_to_control}
#### Mean coverage across replicates
mean.cov.list <- list()
for(sample in unique(sample.info$sample)){
  print(sample)
  ### mean coverage for each position in each gene across replicates
  list1 <- cov.list[[ sample.info$sample_name[sample.info$sample == sample & sample.info$rep == "rep1"] ]]
  list2 <- cov.list[[ sample.info$sample_name[sample.info$sample == sample & sample.info$rep == "rep2"] ]]
  for(m in names(list1)){
    mean.cov.list[[sample]][[m]] <- rowMeans(cbind(list1[[m]], list2[[m]]))
  }
}
if( !is.null(control.sample.names)){
  ### Subtract the mean coverage of each sample by the mean coverage of control sample.
  norm.cov.list <- list()
  for(sample in names(mean.cov.list)){
    if(sample %notin% control.sample){
      print(sample)
      list.drb <- mean.cov.list[[sample]]
      list.ctrl <- mean.cov.list[[control.sample]]
      for(m in names(list1)){
        norm.cov.list[[sample]][[m]] <- pmax(0, list.drb[[m]] - list.ctrl[[m]])
      }
    }
  }
} else{
  norm.cov.list <- mean.cov.list
}
## remove intermediate files
rm(cov.list)
rm(mean.cov.list)
gc()
```

## 5. Smooth read coverage profiles

Similar to [Gregersen et al., 2020](DOI: 10.1038/s41596-019-0262-3), we smooth read coverage profiles using R function smooth.spline. The spar parameter controls the smoothness. We use spar=0.9 for single gene profile.

```{r smooth_profile}
gene.spar=0.9     
norm.spline.list <- list()
for (sample in names(norm.cov.list)){
  ### spline for each sample (mean across replicates) after control normalization
  print(sample)
  norm.spline.list[[sample]] <- sapply(norm.cov.list[[sample]], function(x){pmax(0, round(smooth.spline(1:length(x), x, spar=gene.spar)$y, 3))})
}
saveRDS(norm.cov.list, paste0(output.dir, "norm.cov.list.Rds"))
saveRDS(norm.spline.list, paste0(output.dir, "norm.spline.list.Rds"))
```

## 6. Wave fronts identification

To find the longest distance traveled by RNA Polymerase II (RNAPII) in certain time intervals, we computationally identify the fronts of the elongation waves. 

```{r find_wave_front}
wf <- sapply(norm.spline.list, function(x) {sapply(x, function(i) {wave_front_calling(i)})}) %>% as.data.frame
wf[is.na(wf)] <- 0

# wf_long <-
#   wf %>%
#   rownames_to_column("gene_id") %>%
#   pivot_longer(cols = -gene_id, names_to = "sample", values_to = "wf") %>%
#   mutate(sample = factor(sample, levels = unique(sample.info$sample)))
```

## 7. Elongation rates calculation 

```{r linear_regression}
### order of the wave fronts for each gene
wave.orders <- apply(wf, 1, function(x) {paste0(order(x), collapse = "")})

rate_df <-
  wf %>%
  rownames_to_column("gene_id") %>%
  left_join(gene.gr %>%
              as.data.frame %>%
              mutate(gene_length = width) %>%
              dplyr::select(gene_id, gene_name, gene_type, gene_length)) %>%
  pivot_longer(cols = -c(gene_id, gene_type, gene_name, gene_length), names_to = "DRB", values_to = "wf") %>%
  mutate(wf = as.numeric(wf/1000)) %>%
  mutate(time = as.numeric(sub("minR", "", sub("DRB_", "", DRB)))) %>%
  left_join(as.data.frame(wave.orders) %>%
              rownames_to_column("gene_id"))

### recheck the orders of wave fronts
rate_df %<>%
  group_by(gene_id) %>%
  mutate(order = ifelse((wf[DRB=="DRB_5minR"] < wf[DRB=="DRB_10minR"]) &
                          (wf[DRB=="DRB_10minR"] < wf[DRB=="DRB_15minR"]) &
                          (wf[DRB=="DRB_15minR"] < wf[DRB=="DRB_20minR"]) &
                          (wf[DRB=="DRB_20minR"] != round(gene_length/1000, 1)), "1234",
                        ifelse((wf[DRB=="DRB_5minR"] < wf[DRB=="DRB_10minR"]) &
                                 (wf[DRB=="DRB_10minR"] < wf[DRB=="DRB_15minR"]), "123", "other"))) %>%
  ungroup()

rate_df %<>%
  group_by(gene_id) %>%
  mutate(elongation.rate = ifelse(order == "1234", lm(wf~time)$coef[2],
                                  ifelse(order == "123", lm(wf[DRB!="DRB_20minR"] ~ time[DRB!="DRB_20minR"])$coef[2],
                                         lm(wf~time)$coef[2])),
         adj.r.squared = ifelse(order == "1234", summary(lm(wf~time))$adj.r.squared,
                                ifelse(order == "123", summary(lm(wf[DRB!="DRB_20minR"]~time[DRB!="DRB_20minR"]))$adj.r.squared,
                                       summary(lm(wf~time))$adj.r.squared))) %>%
  ungroup()
```


```{r save_data}
write.table(rate_df, paste0(output.dir, "elongation_rate_tb.txt"), quote=FALSE, sep="\t", row.names=FALSE)
```

```{r}
sessionInfo()
```
